---
title: 'Check convergence of MCMCtree runs'
output:
  html_document:
    df_print: paged
---
See /Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/scripts/0baseml.sh for reference
and https://github.com/mariodosreis/divtime

<b>1. Trace plots.</b> These document the magnitude of the sample drawn (y-axis) at each iteration (x-axis) of the MCMC procedure. Once the chain has identified the stationary distribution of samples, the samples that are drawn will appear to have been randomly sampled from the same region of the y-axis.\
<b>2. Autocorrelation plots.</b> These document the correlation (y-axis) of samples at each step of the chain with previous estimates of that same variable, lagged by some number of iterations (x-axis). <b> Ideally, the autocorrelation declines rapidly, so that it is eventually possible to be confident that the samples from the stationary distribution can be thought of as random, and not reliant on initial values in the chain. </b> If the autocorrelation fails to diminish, a pattern in the sample draws in the trace plot will become more discernible, suggesting that at the very least, the posterior distribution has not been well characterized by the drawn samples.\

```{r echo = F}
#3. Density plot. As its name implies, this plot is a summary of the sampled values that define the stationary distribution of values, which approximates the posterior distribution of interest. The peak of the density (the maximum a posteriori, or MAP, estimate) is the mode of the distribution, which is the value with the most support from the data and the specified prior. Other posterior summaries, such as the posterior mean or posterior median, are also easily obtained. Kernel density estimation is used to smooth over the samples and produce an estimate of the posterior distribution. It should be noted that the density plot is not a formal means of diagnosing model convergence, as are the autocorrelation and trace plots. However, unexpected peaks or strange shapes in the posterior density can be a sign of poor model convergence.

```

```{r echo=FALSE}
mcmc1<-read.table("/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/mcmc.run3.txt",T)
mcmc2<-read.table("/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/mcmc.run4.txt",T)
```

Each data frame contains 14 columns:
MCMC generation number, 10 node ages (divergence times), 1 mean mutation rates, 1 rate drift coefficients, and sample log-likelihood values
```{r}
names(mcmc1)
```

To check for convergence of the MCMC runs, we calculate the posterior means of times for each run, and plot them against each other (a). Good convergence is indicated when the points fall on the y = x line. Trace plots (b,c) are useful to visualise the MCMC and split problems. In a good mcmc run, these should look like "hairy caterpillars".
If you see a persistent trend in the trace, the MCMC likely did not converge and we need a longer burn-in.\
Figure d shows the smoothed histograms (calculated using density in R) for t_n11
for the two runs. Notice that the two histograms are slightly different. As the ESS
becomes larger, histograms for different runs will converge in shape until becoming
indistinguishable. If you see large discrepancies between histograms, that may indicate
serious problems with the MCMC, such as lack of convergence due to short burn-in, or
the MCMC getting stuck in different modes of a multi-modal posterior.\

```{r echo=FALSE, fig.height=2, fig.width=2}
t.mean1 <- apply(mcmc1[,2:11], 2, mean) * 100
t.mean2 <- apply(mcmc2[,2:11], 2, mean) * 100

#par(mfrow=c(2,2))
#dev.off()

plot(t.mean1, t.mean2, main="Posterior times, run 1 vs. run 2")
abline(0, 1)
#dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/mcmc.r3.r4.correlation.pdf")

plot(mcmc1$t_n19, ty='l', main="b) trace of t_n19")
plot(mcmc1$t_n12, ty='l', main="c) trace of t_n12")

plot(density(mcmc1$t_n12), main="d) histogram of t_n12, r 1 vs. r2")
lines(density(mcmc2$t_n12), lty=2)



```

Plotting autocorrelation of the different variables that are included in the model. Ideally, the autocorrelation quickly decreases and reaches 0. 

```{R echo=FALSE}
library(plotMCMC)
```

Here is the autocorrelation for the first mcmc run.\

```{r echo=F}
par(mai=c(.4,.5,.4,.2)) # this changes the margin sizes
  plotAuto(mcmc1[,c(2,6,8,10,12,13)])
  
```

```{r eval=FALSE, include=FALSE}
#And for the second mcmc run.
par(mai=c(.4,.5,.4,.2)) # this changes the margin sizes
plotAuto(mcmc2[,2:13])
```


We can calculate the effective sample sizes (ESS) of the parameters (using the coda package).

```{R echo=FALSE}
mean.mcmc <- apply(mcmc1[,-1], 2, mean)
ess.mcmc <- apply(mcmc1[,-1], 2, coda::effectiveSize)
var.mcmc <- apply(mcmc1[,-1], 2, var)
se.mcmc <- sqrt(var.mcmc / ess.mcmc)
cbind(mean.mcmc, ess.mcmc, var.mcmc, se.mcmc)

# oldpar <- par()
# mai <- oldpar$mai; mai[3] <- .35; par(mai = mai)
# mai <- oldpar$mai; mai[1] <- .35; mai[2] <- .35; mai[3] <- .45; par(mai = mai)

mean.mcmc <- apply(mcmc2[,-1], 2, mean)
ess.mcmc <- apply(mcmc2[,-1], 2, coda::effectiveSize)
var.mcmc <- apply(mcmc2[,-1], 2, var)
se.mcmc <- sqrt(var.mcmc / ess.mcmc)
cbind(mean.mcmc, ess.mcmc, var.mcmc, se.mcmc)

```
The ESS should ideally be >1000.


# Prior

To see if the priors fit the sampling by the MCMC chain relatively well, we can check the mcmc runs where we only sampled from the prior.
```{R echo =F}
# ###############################################
# PRIOR:
# ###############################################
mcmc1.p <- read.table("/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/p.mcmc.run3.txt",T)

mcmc2.p <- read.table("/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/p.mcmc.run4.txt",T)
```

Check that run p1 and p2 reach the same posterior dates for each variable. 
```{r fig.height=3, fig.width=3}
p.mean1 <- apply(mcmc1[,2:11], 2, mean) * 100
p.mean2 <- apply(mcmc2[,2:11], 2, mean) * 100
plot(p.mean1, p.mean2); abline(0, 1)


```

Compare the prior and the posterior for each variable. \
```{r echo =F }
  par(mai=c(.2,.3,.2,.1)) # this changes the margin sizes
  par(mar=c(2,2,1,1)) # this changes the margin sizes
  par(mfcol=c(3,5))
  for(i in 12:1) {
    dpr <- density(mcmc1.p[,i+1], adj=.1) # prior
    dPr <- density(mcmc1[,i+1], adj=.1)   # Posterior
    xl <- range(c(dpr$x, dPr$x))
    yl <- range(c(dpr$y, dPr$y))
    plot(dpr, main=paste("t_n",i+10,sep=""), xlab="", ylab="", las=1, xlim=xl, ylim=yl, col="darkgrey")
    lines(dPr, col="black")
  }
    dpr <- density(mcmc1.p$sigma2, adj=.1) # prior
    dPr <- density(mcmc1$sigma2, adj=.1)   # Posterior
    
    # Gamma distribution with shape alpha = 5.5 and scale b = 1
    #dDg <- density(log(rgamma(10000,shape=5.5,scale=1),10),adj=.1) 
    #dDg <- density(rgamma(100000,shape=1,scale=0.4),adj=.1) 
    
    xl <- range(c(dpr$x, dPr$x))
    yl <- range(c(dpr$y, dPr$y))
    plot(dpr, main=paste("sigma2",sep=""), xlab="", ylab="", las=1, xlim=xl, ylim=yl, col="darkgrey")
    lines(dPr, col="black")
    #lines(dDg, col="darkgrey",lty=2)
    #dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/mcmc.r3.priorVSposterior.pdf")
```

This looks ok and we don't see substantial incompatibilities between prior and posterior distribution.

\
Now, we plot the tree with the appropriate error bars.
 
Prepare plotting
```{r include=FALSE}
# prepare plotting
library(phytools)
library(plotrix)
#install.packages("devtools")
library(devtools)
#install_github("PuttickMacroevolution/MCMCTreeR")
library(MCMCTreeR)
```

#Run3
```{r fig.height=5, fig.width=12, echo =F}
mcmctree<-readMCMCTree("/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/FigTree.run3.tre")
tree<-mcmctree$apePhy
CI<-mcmctree$nodeAges
plotTree(tree,xlim=c(.4,-.1),direction="leftwards",
    mar=c(4.1,1.1,1.1,.1),ftype="i")

abline(v=seq(0,.4,by=10),lty="dashed",
    col=make.transparent("grey",0.5))
axis(1,at=seq(0,.4,by=.1),line=2,labels=c("present","10 MYA","20 MYA","30 MYA","40 MYA"))
obj<-get("last_plot.phylo",envir=.PlotPhyloEnv)
#obj
par(xpd=NA)

for(i in 1:tree$Nnode+Ntip(tree)){
    lines(x=c(CI[i-Ntip(tree),2],CI[i-Ntip(tree),3]),
        y=rep(obj$yy[i],2),lwd=4,lend=0,
        col=make.transparent("blue",0.4))
    points(obj$xx[1:tree$Nnode+Ntip(tree)],
        obj$yy[1:tree$Nnode+Ntip(tree)],pch=19,col="blue",cex=1)
    segments(x0 = CI[i-Ntip(tree),1],x1 = CI[i-Ntip(tree),1],y0=min(obj$yy)-.6,y1=max(obj$yy)+0.6,lty=2,col=rgb(0,0,0,.4))
    text(round(CI[i-Ntip(tree),1],4)*100,x = CI[i-Ntip(tree),1]+.015,y=obj$yy[i],pos = 3)
    text(round(CI[i-Ntip(tree),1],4)*100,x = CI[i-Ntip(tree),1],y=min(obj$yy)-1.5,pos = 3,srt=45,cex=.5)
}
par(xpd=T)
CI
#dev.print(pdf,        # copies the plot to a the PDF file
#         "/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/FigTree.run3.pdf")
```

#Run4
```{r fig.height=5, fig.width=12, echo =F}
mcmctree<-readMCMCTree("/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/FigTree.run4.tre")
tree<-mcmctree$apePhy
CI<-mcmctree$nodeAges
plotTree(tree,xlim=c(.4,-.1),direction="leftwards",
    mar=c(4.1,1.1,1.1,.1),ftype="i")

abline(v=seq(0,.4,by=10),lty="dashed",
    col=make.transparent("grey",0.5))
axis(1,at=seq(0,.4,by=.1),line=2,labels=c("present","10 MYA","20 MYA","30 MYA","40 MYA"))
obj<-get("last_plot.phylo",envir=.PlotPhyloEnv)
#obj
par(xpd=NA)

for(i in 1:tree$Nnode+Ntip(tree)){
    lines(x=c(CI[i-Ntip(tree),2],CI[i-Ntip(tree),3]),
        y=rep(obj$yy[i],2),lwd=4,lend=0,
        col=make.transparent("blue",0.4))
    points(obj$xx[1:tree$Nnode+Ntip(tree)],
        obj$yy[1:tree$Nnode+Ntip(tree)],pch=19,col="blue",cex=1)
    segments(x0 = CI[i-Ntip(tree),1],x1 = CI[i-Ntip(tree),1],y0=min(obj$yy)-.6,y1=max(obj$yy)+0.6,lty=2,col=rgb(0,0,0,.4))
    text(round(CI[i-Ntip(tree),1],4)*100,x = CI[i-Ntip(tree),1]+.015,y=obj$yy[i],pos = 3)
    text(round(CI[i-Ntip(tree),1],4)*100,x = CI[i-Ntip(tree),1],y=min(obj$yy)-1.5,pos = 3,srt=45,cex=.5)
}
par(xpd=T)
CI

#dev.print(pdf,        # copies the plot to a the PDF file
#          "/Users/lukas/sciebo/inquilineGenomics18/phylogeny/TreeDating/MCMCtree/results/FigTree.run4.pdf")
  
```