---
title: "Inquiline Genomics: GR gene tree analysis and plot"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: true
    number_sections: true
    df_print: paged
  pdf_document: default
---
<STYLE TYPE="text/css">
<!--
  td{
    font-family: Arial; 
    font-size: 8pt;
    padding:0px;
    cellpadding="0";
    cellspacing="0"
  }
  th {
    font-family: Arial; 
    font-size: 8pt;
    height: 20px;
    font-weight: bold;
    text-align: right;
    background-color: #ccccff;
  }
  table { 
    border-spacing: 0px;
    border-collapse: collapse;
  }
--->
</STYLE>

# Load Environment
## Load libraries

```{r include=FALSE}
######################
# Load packages
######################
source("~/sciebo/librarySchrader.R")
library("ape")
library("Biostrings")
library("ggplot2")
library("ggtree")
library("phytools")
library("treeman")
library(diversitree)
library(phytools)
library(geiger)
library(cowplot)
library(RColorBrewer)
library(phylobase)
library(gplots)
library(randomcoloR)
library(data.table)
```

## Load tree and midpoint root it
```{r}
#https://joey711.github.io/phyloseq/plot_tree-examples.html
############################################
# Load rerooted FastTree tree
############################################
t1<-read.tree("/Users/lukas/sciebo/inquilineGenomics18/geneFamilies/specificFamilies/GRevolution/GRs.complete.fast.tre")
t1<-midpoint.root(t1)
tree <- as(t1, 'TreeMan')
```

## Load tsv with information for each gene
Focus on domain-containing ("7_TM7") and complete genes (="M*") 

```{r}
############################################
# Load subfam assignments
############################################
subfam<-read.table("/Users/lukas/sciebo/inquilineGenomics18/geneFamilies/specificFamilies/GRevolution/GRs.details.tsv",sep="\t")
subfam<-subset(subfam,V3=="M*" & V2=="7_TM7")

```


# Process tree data
## Generate clades of min size 4
```{r}

############################################
# Generate clades
############################################
cladeSizeCutoff<-4
cladesTree<-getCladesofSize(t1,cladeSizeCutoff)

# generate overview dataframe
tipList<-lapply(cladesTree, function(x) x$tip.label)
names(tipList)<-paste("",sprintf("%03d", 1:length(tipList)),sep="")
tipDF<-stack(tipList)
cladeSizes<-lapply(tipList, length)
cladeSizesTab<-do.call(rbind,cladeSizes)
plot(cladeSizesTab,cex=.8,pch=19,col=addalpha(goodCols[1],.8))
abline(h=cladeSizeCutoff,col="red",lwd=2,lty=2)
cladeSummary<-lapply(tipList, function(x) summary(as.factor(substr(x,1,4))))
cladeTable<-plyr::ldply(cladeSummary, rbind)
cladeTable[is.na(cladeTable)] <- 0
dim(cladeTable)

```


## Identify complete and fragmented genes
```{r}
tipDF2<-merge(tipDF,subfam,by.x="values",by.y="V1")
colnames(tipDF2)<-c("gene","clade","domain","completeness","V4","model","source","original","pos","seq")
tipDF2$status<-as.character(tipDF2$completeness)
tipDF2$status[grep("M\\*$",tipDF2$completeness)]<-"0"
tipDF2$status[grep("M[^\\*]$",tipDF2$completeness)]<-"CTE"
tipDF2$status[grep("^[^M]\\*$",tipDF2$completeness)]<-"NTE"
tipDF2$status[grep("^[^M][^\\*]$",tipDF2$completeness)]<-"NC"
tipDF2<-  tipDF2[,c(1:4,11)]

```


## Add subfamily metadata to tree
```{r}
# check all group labels
#levels(attributes(t1)$group) 
tipDF2$species<-substr(1,4,x=as.character(tipDF2$gene))
tipDF2$type<-"Atta"
tipDF2$type[grep("Ahey|Aech",tipDF2$species,perl=T)]<-"Acromyrmex"
tipDF2$type[grep("Acha|Ains|Parg",tipDF2$species,perl=T)]<-"inquiline"
tipDF2$type<-as.factor(tipDF2$type)
subfamList<-unstack(tipDF2[,c("gene","type")])
t1 <- ggtree::groupOTU(t1, subfamList, overlap='abandon')

levels(attributes(t1)$group)

```

## retrieve genes according to clade membership from tree object
```{r}
tipDF2list<-lapply(cladesTree, function(x) subset(tipDF2,gene %in% x$tip.label))
names(cladesTree)<-unlist(lapply(tipDF2list,function(x) x$clade[1]))
```

# Generate trait data for genes & clades
```{r}
############################################
# Generate traits
############################################

smallerCladesTrait<-rep(0,length(t1$tip.label))
names(smallerCladesTrait)<-t1$tip.label

## Tier 1: clade evolution
## level 1: All clades decreased in one parasite
# level 1a: All clades decreased in Parg relative to Ahey
smallerClades1a<-as.numeric(row.names(subset(cladeTable,(Parg<Ahey   & Ahey <= Aech))))
smallerClades1aTree<-cladesTree[smallerClades1a]
smallerClades1aTips<-unlist(lapply(smallerClades1aTree, function(x) x$tip.label))

# level 1b: All decreased in Acha relative to Ahey 
smallerClades1b<-as.numeric(row.names(subset(cladeTable,(Ahey <= Aech & Acha<Ahey))))
smallerClades1bTree<-cladesTree[smallerClades1b]
smallerClades1bTips<-unlist(lapply(smallerClades1bTree, function(x) x$tip.label))

# level 1c: All decreased in Ains relative to Aech
smallerClades1c<-as.numeric(row.names(subset(cladeTable,(Ains<Aech   & Aech <= Ahey))))
smallerClades1cTree<-cladesTree[smallerClades1c]
smallerClades1cTips<-unlist(lapply(smallerClades1cTree, function(x) x$tip.label))

# level 2: Decreases in two parasites
# level 2a: All decreased in Ains & Parg
smallerClades2a<-as.numeric(row.names(subset(cladeTable,(Parg<Ahey) & (Ains<Aech) & (Acha>=Ahey))))
smallerClades2aTree<-cladesTree[smallerClades2a]
smallerClades2aTips<-unlist(lapply(smallerClades2aTree, function(x) x$tip.label))

# level 2b: All decreased in Ains & Acha
smallerClades2b<-as.numeric(row.names(subset(cladeTable,(Acha<Ahey) & (Ains<Aech) & (Parg>=Ahey))))
smallerClades2bTree<-cladesTree[smallerClades2b]
smallerClades2bTips<-unlist(lapply(smallerClades2bTree, function(x) x$tip.label))

# level 2c: All decreased in Parg & Acha (& Ahey >=Aech)
smallerClades2c<-as.numeric(row.names(subset(cladeTable,(Acha<Ahey) & (Parg<Ahey) & (Ahey<=Aech))))
smallerClades2cTree<-cladesTree[smallerClades2c]
smallerClades2cTips<-unlist(lapply(smallerClades2cTree, function(x) x$tip.label))

# level 3:  All clades decreased in all 3 parasites
smallerClades3<-as.numeric(row.names(subset(cladeTable,Parg<Ahey & Acha<Ahey & Ains<Aech)))
smallerClades3Tree<-cladesTree[smallerClades3]
smallerClades3Tips<-unlist(lapply(smallerClades3Tree, function(x) x$tip.label))

# finalize trait data frame
smallerCladesTrait[smallerClades1aTips]<-"Parg"
smallerCladesTrait[smallerClades1bTips]<-"Acha"
smallerCladesTrait[smallerClades1cTips]<-"Ains"
smallerCladesTrait[smallerClades2aTips]<-"Parg/Ains"
smallerCladesTrait[smallerClades2bTips]<-"Acha/Ains"
smallerCladesTrait[smallerClades2cTips]<-"Parg/Acha"
smallerCladesTrait[smallerClades3Tips]<-"Parg/Ains/Acha"

smallerCladesTrait[smallerClades1aTips]<-"1"
smallerCladesTrait[smallerClades1bTips]<-"1"
smallerCladesTrait[smallerClades1cTips]<-"1"
smallerCladesTrait[smallerClades2aTips]<-"2"
smallerCladesTrait[smallerClades2bTips]<-"2"
smallerCladesTrait[smallerClades2cTips]<-"2"
smallerCladesTrait[smallerClades3Tips]<-"3"


smallerCladesTrait<-smallerCladesTrait[t1$tip.label]
smallerCladesTrait<-as.factor(smallerCladesTrait)

# level 1:  All complete gene models
subfamTrait<-as.factor(tipDF2$type)
names(subfamTrait)<-tipDF2$gene
#subfamTrait<-subfamTrait[t1$tip.label]
```


## Plot heatmap of gene copies in clades vs species
```{r}
cladeMatrixLC<-cladeTable[,c("Aech","Ains","Parg","Ahey","Acha","Acol","Acep")]
cladeMatrixLC2<-cladeMatrixLC[apply(cladeMatrixLC, 1, var) != 0, ]
color.palette  <- colorRampPalette(c(goodCols[5],goodCols[1]))(n=10)
heatmap.2(as.matrix(cladeMatrixLC2), col = color.palette,scale="none",trace="none",key.title = "",key.par=list(mar=c(3.5,1,.1,.1)),key.xlab = "copies in clade")
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/geneFamilies/specificFamilies/GRevolution/GRs.heatmap.pdf",width=4,height=7)

```

## create color vectors for all trait data
```{r}
n <- length(levels(subfamTrait))
n2 <- length(levels(smallerCladesTrait))
n4 <- length(levels(attributes(t1)$group))

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]

# subfamilies or clades
col_vector <- c(addalpha(goodCols[1],.3),addalpha(goodCols[2],.3),addalpha(goodCols[3],.6))

# parasite losses
fc<-colorRampPalette(c("#FFCCCC", goodCols[8]))
col_vector2 <- c("gray90",fc(n2-1))


# branch coloring
col_vector4 <- rep("black",(n4+1))

```

## prepare tree with outside heatmaps
```{r}
### 2 outline circles
circ <- ggtree(t1,size=.2, layout = "circular",ladderize = T,right = T,mapping = aes(color = group)) + scale_color_manual(name = 'group',values = col_vector4) 

# 1st tier: species
p1 <- gheatmap(circ, data.frame(subfamTrait), offset=0.31, width=.02,
               colnames_angle=90, colnames_offset_y = 0.5,color = NA,colnames = F)
# 2nd tier: Clades changing in parasites
p2 <- gheatmap(p1, data.frame(smallerCladesTrait), offset=.23, width=.02,
               colnames_angle=90, colnames_offset_y = .25,color = NA,colnames = F)

col <- c(col_vector, col_vector2)
names(col) = c(levels(subfamTrait), levels(smallerCladesTrait))

# changed to 2nd tier
pp <- p2 + scale_fill_manual(values=col)

p1x <- p1 + scale_fill_manual(values=col)
p2x <- gheatmap(circ, data.frame(smallerCladesTrait), offset=.15, width=.02,color = NA) +
    scale_fill_manual(values=col) 

leg1 <- get_legend(p1x)
leg2 <- get_legend(p2x)

pp <- pp + theme(legend.position="none") 
```

## plot legend
```{r}

plot_grid(leg1,leg2, ncol=4)
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/geneFamilies/specificFamilies/GRevolution/GRs.labeled.legend.pdf",width=10,height=10)

```

## add clade info to clades in ggtree object
```{r}
pp2 <- pp %<+% tipDF2
tree_dt <- data.table(pp2$data)


# select only the tip labels and order by coord y
tree_dt <- tree_dt[isTip == TRUE][order(y)]

# Make table with y cords for each group in consecutive order; 
# this helps for drawing & labeling segments.
# Note the usage of "rleid" function, which is a grouping ID generator,
# needed because consecutive rows of an identical reoccurring group must form a unique group.
coord_groups <- tree_dt[, .(y1 = y[1],
                            y2 = y[.N],
                            angle = mean(angle),
                            n = .N), # optional - helps with counting
                        by = .(clade, 
                               id_gr = rleid(clade, 
                                             prefix = "cld"))]
coord_groups


# Compute the middle y - will be used for placing the group label;
# similarly the mean angle was computed above already. 
coord_groups[, y_mid := rowMeans(.SD), .SDcols = c("y1", "y2")]

# For one-record groups where y1=y2, adjust their y coordinates so that a segment gets drawn;
# If not, no segment get drawn for such cases.
# To force a segment add and subtract a small amount (try & error until seems ok). 
# Prefer a smaller values since with bigger ones you risk to exaggerate the segments.
coord_groups[, y1_adj := ifelse(y1 == y2, y1 - 0.1, y1)]
coord_groups[, y2_adj := ifelse(y1 == y2, y2 + 0.1, y2)]

# Labels need angle adjustment for cases between 90 and 270 dg
coord_groups[, angle_adj := ifelse(angle %between% c(90, 180), 
                                   yes = angle + 180,
                                   no = ifelse(angle > 180 & angle <= 270,
                                               yes = angle - 180,
                                               no = angle))]

# Labels with angles between 90 and 270 dg
# need change of horizontal adjustment argument from 0 to 1.
coord_groups[, hjust_adj := ifelse(angle %between% c(90, 270), yes = 1L, no = 0L)]

# if needed, coloring could be binary 
coord_groups[, col := ifelse(.I%%2, 0.35, 1)]
coord_groups


# Define variable to control x coordinate of segments & labels
my_x <- max(tree_dt$x) + 0.3

tree_labeled <- 
  pp2 + 
  # Add line segments for each group.
  geom_segment(data = coord_groups,
               aes(x = my_x, 
                   y = y1_adj, 
                   xend = my_x, 
                   yend = y2_adj),
               color = rgb(.5,.5,.5,.5),
               lineend = "butt",
               size = 1.2) +
  # Add text group labels at the middle of each segment.
  geom_text(data = coord_groups,
            aes(x = my_x+0.01,
                y = y_mid,
                angle = angle_adj,
                hjust = hjust_adj,
                label = clade),
            vjust = 0.5, 
            size  = 4,
            nudge_x = 0.06, # Offsetting label from its default x coordinate.
            color = "black")
```

# Plot final tree
```{r}
tree_labeled+ geom_tiplab2(size=1, aes(angle=angle),align=T,linesize=.1,linetype = "dashed", color="gray90") + geom_tiplab2(size=1, aes(angle=angle),align=T,linesize=0,linetype = NA, color="gray40")
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/geneFamilies/specificFamilies/GRevolution/GRs.labeled.tree.pdf",width=7.5,height=7.5)

```

# Save relevant objects
```{r}
dfSC<-data.frame(smallerCladesTrait)
tipDF4<-merge(tipDF2,dfSC,by.x="gene",by.y="row.names")

# save some things
save(cladesTree,file="/Users/lukas/sciebo/inquilineGenomics18/geneFamilies/specificFamilies/GRevolution/GRs.labeled.tree.Robject")
write.table(tipDF4,"/Users/lukas/sciebo/inquilineGenomics18/geneFamilies/specificFamilies/GRevolution/GRs.tipsDF4.tsv",quote=F,sep="\t",row.names = F)
write.table(tree_dt,"/Users/lukas/sciebo/inquilineGenomics18/geneFamilies/specificFamilies/GRevolution/GRs.tree_dt.tsv",quote=F,sep="\t",row.names = F)
write.table(cladeTable,"/Users/lukas/sciebo/inquilineGenomics18/geneFamilies/specificFamilies/GRevolution/GRs.cladeTable.tsv",sep="\t",quote = F,row.names=F)
```

