---
title: "Inquiline Genomics: GO enrichment analysis of RELAX results"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: true
    number_sections: true
    df_print: paged
  pdf_document: default
---
<STYLE TYPE="text/css">
<!--
  td{
    font-family: Arial; 
    font-size: 8pt;
    padding:0px;
    cellpadding="0";
    cellspacing="0"
  }
  th {
    font-family: Arial; 
    font-size: 8pt;
    height: 20px;
    font-weight: bold;
    text-align: right;
    background-color: #ccccff;
  }
  table { 
    border-spacing: 0px;
    border-collapse: collapse;
  }
--->
</STYLE>

# Load Environment
## Load libraries

```{r include=FALSE}
###############
# Load packages
###############

source("~/sciebo/librarySchrader.R")
library("ape")
library("Biostrings")
library("ggplot2")
library("ggtree")
library("phytools")
library("treeman")
library(diversitree)
library(phytools)
library(geiger)
library(cowplot)
library(RColorBrewer)
library(gplots)
library(randomcoloR)
library(data.table)
library(viridis)
library(ggplot2)
library(dplyr)
library(pvclust)
library(jmuOutlier)
```
## Load data
```{r}
# load cladesTree object
load("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/cladesTree.Robject")
# load tree_dt object
load("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/tree_dt.Robject")
# color table
colorTableSF<-read.csv("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/colorTableSF.tsv",sep="\t",header=T,stringsAsFactors = F)

tipDF4<-read.table("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/tipsDF4.tsv",T,sep="\t")
allRELAX<-read.table("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/OR.SOS/results/allRELAX.tsv",T)
allRELAX$klog2<-log(allRELAX$k,2)

```

## Generate subtrees
```{r}
tipDF4$species<-as.factor(substr(tipDF4$gene,1,4))

tipDF4$parasite<-NA
tipDF4$parasite[grep("Acha",tipDF4$species)]<-"Acha"
tipDF4$parasite[grep("Ains",tipDF4$species)]<-"Ains"
tipDF4$parasite[grep("Parg",tipDF4$species)]<-"Parg"


targetClades<-unique(tipDF4$clade)
tipDF4subset<-tipDF4


subsetTrees<-cladesTree[targetClades] 

ggSubsetTrees<-list()
ggST<-list()
for (i in 1:length(subsetTrees)){
  subtit<-paste("group=", tipDF4[tipDF4$clade==targetClades[i],5][1],"SF=",tipDF4[tipDF4$clade==targetClades[i],3][1],sep=" ")
  ggSubsetTrees[[i]]<-ggtree(subsetTrees[[i]],ladderize = T,right = T)
  infoDF<-subset(tipDF4subset,gene %in% subsetTrees[[i]]$tip.label)
  ggSubsetTrees[[i]]<-ggSubsetTrees[[i]]%<+% infoDF
  ggST[[i]]<-ggSubsetTrees[[i]] + geom_tiplab(cex=2,aes(fill = factor(parasite)),
                color = "black", # color for label font
                geom = "label",  # labels not text
                label.padding = unit(0.15, "lines"), # amount of padding around the labels
                label.size = 0) + 
             scale_fill_manual(na.value = NA, values=c(Ains =goodCols[2], Parg = goodCols[5],Acha =goodCols[1])) +
             #theme(legend.position = "bottom", 
                #legend.title = element_blank(), # no title
                #legend.key = element_blank()) +
             ggtitle(paste("Clade",targetClades[i],sep=" "),subtitle =subtit) + xlim(0, .3)+
             theme(plot.title = element_text(size = 12, face = "bold"),plot.subtitle = element_text(size = 6))
  }

```
## create plots for each clade
```{r eval=FALSE, fig.height=10, fig.width=10, include=FALSE}
for (p in levels(tipDF4$smallerCladesTrait)){
  folderName<-gsub("/",".",p)
  #dir.create(path = paste("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/CladePlots/",folderName,sep=""))
}

for (i in 1:length(ggST)){
  cladetype<-gsub("/",".",tipDF4[tipDF4$clade==targetClades[i],5][1])
  subfam<-tipDF4[tipDF4$clade==targetClades[i],3][1]
  #pdf(paste("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/CladePlots/",cladetype,"/",subfam,"-Clade-",targetClades[i],".pdf",sep=""),width=3,height=5)
  #print(ggST[[i]])
  #dev.off()
}

```
## Assign clades
```{r eval=FALSE, include=FALSE}
for (p in levels(tipDF4$smallerCladesTrait)){
  folderName<-gsub("/",".",p)
  #dir.create(path = paste("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/clades/",folderName,sep=""))
}

for (i in 1:length(subsetTrees)){
cladetype<-gsub("/",".",tipDF4[tipDF4$clade==targetClades[i],5][1])
subfam<-tipDF4[tipDF4$clade==targetClades[i],3][1]
write.tree(subsetTrees[[i]],paste("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/clades/",cladetype,"/",subfam,"-Clade-",names(subsetTrees)[i],".tre",sep=""))
tipDF4.clade<- subset(tipDF4,gene %in% subsetTrees[[i]]$tip.label)

write.table(tipDF4.clade,paste("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/clades/",cladetype,"/",subfam,"-Clade-",names(subsetTrees)[i],".tsv",sep=""),sep="\t",quote = F,row.names = F)
}

```

## create annotation tables for subfamilies
```{r}
tipDF4$annotation<-"complete"
tipDF4$annotation[grep("CTE|NTE|NC",tipDF4$gene,perl=T)]<-"fragment"
tipDF4$annotation[grep("dH$",tipDF4$gene,perl=T)]<-"dH"

tipList<-split(tipDF4,f=list(tipDF4$subfamily,tipDF4$annotation))
tipList2<-(lapply(tipList, function(x) summary(x$species)))

annotationTable<-do.call(rbind,tipList2)
annotationTable<-annotationTable[sort(row.names(annotationTable)),]
annotationTable<-annotationTable[rowSums(annotationTable)!=0,]
annotationTable
annotationTableTmp<-cbind(row.names(annotationTable),annotationTable)
colnames(annotationTableTmp)[1]<-"subfamily"
write.table(annotationTableTmp,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/annotationTable.tsv",sep="\t",quote = F,row.names = F)
tipListAll<-split(tipDF4,f=list(tipDF4$subfamily))
tipListAll2<-(lapply(tipListAll, function(x) summary(x$species)))

annotationTable2<-do.call(rbind,tipListAll2)
annotationTable2<-annotationTable2[sort(row.names(annotationTable2)),]
annotationTable2<-annotationTable2[rowSums(annotationTable2)!=0,]
annotationTable2
write.table(annotationTable2,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/annotationTable2.tsv",sep="\t",quote = F,row.names = T)


```
## Plot clustering & heatmap
```{r}

annotationTable3<-annotationTable2[apply(annotationTable2, 1, var)> 0,]
#color.palette  <- colorRampPalette(c(goodCols[1],goodCols[8]))(n=50)
color.palette<-viridis(50)
colSideHM<-colorTableSF$col
names(colSideHM)<-colorTableSF$sf
tmp<-heatmap.2(t(as.matrix(annotationTable3)), col = color.palette,scale = "col",trace="none")
heatmap.2(t(as.matrix(annotationTable3)), col = color.palette,scale = "col",trace="none",ColSideColors=colSideHM[names(tmp$colMeans)],offsetCol =-32.9,cexCol=.6,adjCol=c(.5,0),srtCol=0,    key=FALSE)
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/OR.heatmap.pdf",width=4,height=4)
heatmap.2(t(as.matrix(annotationTable3)), col = color.palette,scale = "col",trace="none",ColSideColors=colSideHM[names(tmp$colMeans)],offsetCol =-100.5,cexCol=.6,adjCol=c(.5,0),srtCol=0)
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/OR.heatmap.large.pdf",width=10,height=10)
pvc<-pvclust((as.matrix(annotationTable3)),  method.hclust="complete", method.dist="euclidean", nboot = 10000,parallel=T)
plot(pvc)
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/OR.heatmap.PVCLUST.pdf")
```


# Analyse RELAX results
## color Table
```{r}
allRELAX<-merge(allRELAX,colorTableSF,by.x="sf",by.y="sf",all.x=T)
allRELAX$klog2[!is.finite(allRELAX$klog2)]<-(-80)
```

## Calculate bg omega, omega3 and log(k)
```{r}
# Background Omega
backgroundOmega<-data.frame(cbind(allRELAX$ToMean,allRELAX$RoMean))
colnames(backgroundOmega)<-c("T","R")
backgroundOmegaS<-stack(backgroundOmega)
pB <- ggplot(backgroundOmegaS, aes(x=ind, y=values,fill=ind)) + 
  geom_violin(lwd=0.1) + 
  theme(axis.text.x = element_text(angle = 0, hjust = 0)) + 
  scale_y_continuous(name =expression(omega["bg"] )) +
  scale_x_discrete(name ="") +
  scale_fill_manual(values=goodCols[c(3,1)]) +  
  guides(fill=FALSE)+
  geom_boxplot(width=0.1,outlier.alpha = 0,fill=rgb(1,1,1,.8))

# Omega3 rates
Omega3<-data.frame(cbind(allRELAX$To3,allRELAX$Ro3))
colnames(Omega3)<-c("T","R")
Omega3S<-stack(Omega3)
p3 <- ggplot(Omega3S, aes(x=ind, y=log(values,10),fill=ind)) + 
  geom_violin(lwd=0.1)+
  scale_y_continuous(name =expression(paste("lg(",omega["3"],")",sep="")) )+
  scale_x_discrete(name ="") +
  scale_fill_manual(values=goodCols[c(3,1)]) +  
  guides(fill=FALSE)+
  geom_boxplot(width=0.1,outlier.alpha = 0,fill=rgb(1,1,1,.8))

# K distribution
kvals<-log(allRELAX$k,2)
allRELAX$klog2<-log(allRELAX$k,2)
pK1 <- ggplot(allRELAX, aes(x="",y=klog2,fill=goodCols[2])) + 
  geom_violin(col=goodCols[2]) +
  scale_fill_manual(values=goodCols[c(2)]) +
  scale_x_discrete(name ="") +
  guides(fill=FALSE)+
  geom_boxplot(width=0.1,outlier.alpha = 0,fill="white")+
  scale_y_continuous(name=expression(paste(log["2"],"(k)",sep="")) ,limits=c(-40,40))+
  geom_hline(yintercept=0, linetype="dashed", color = "red")

pK2 <- ggplot(allRELAX, aes(x="",y=klog2,fill=goodCols[2])) + 
  geom_violin(col=goodCols[2]) +
  scale_fill_manual(values=goodCols[c(2)]) +
  scale_x_discrete(name ="") +
  guides(fill=FALSE)+
  geom_boxplot(width=0.1,outlier.alpha = 0,fill="white")+
  scale_y_continuous(name="",limits=c(-2.5,2.5))+
  geom_hline(yintercept=0, linetype="dashed", color = "red")+
  theme(axis.text.y = element_text(size = 6))

pK<-ggdraw() +
  draw_plot(pK1 + theme(legend.justification = "bottom"), 0, 0, 1, 1) +
  draw_plot(pK2 + scale_color_viridis(discrete = TRUE) + 
              theme(legend.justification = "top"), 0.5, 0.52, 0.5, 0.4)
plot_grid(pB,p3,pK,ncol=3,labels = c("A","B","C"))

```

## Permutation tests
```{r include=FALSE}
backgroundPT<-perm.test(allRELAX$ToMean,allRELAX$RoMean,stat=median,paired=T)
print(backgroundPT)
omega3PT<-perm.test(allRELAX$To3[!is.na(allRELAX$To3) & !is.na(allRELAX$Ro3)],allRELAX$Ro3[!is.na(allRELAX$To3) & !is.na(allRELAX$Ro3)],stat=median,paired=T)
print(omega3PT)
median(allRELAX$klog2)
kPT<-perm.test(allRELAX$klog2,mu = 0,stat=median)
print(kPT)
```


## Create plot with p-values
```{r echo=FALSE}

pB<-pB+scale_x_discrete(name =paste("perm. test: T > R; p=",ifelse(backgroundPT$p.value==0,"1e-16",backgroundPT$p.value),"\n\nn=20000",sep="")) + theme(axis.title.x = element_text(size = 10))
p3<-p3+scale_x_discrete(name =paste("perm. test: T < R; p=",ifelse(omega3PT$p.value==0,"1e-16",omega3PT$p.value),"\n\nn=20000",sep="")) + theme(axis.title.x = element_text(size = 10))
pK1<-pK1+scale_x_discrete(name =paste("perm. test: log2(k)<0;\n p=",ifelse(kPT$p.value==0,"1e-16",round(kPT$p.value,3)),"\nn=20000",sep="")) + theme(axis.title.x = element_text(size = 10))

library(viridis)
pK<-ggdraw() +
  draw_plot(pK1 + theme(legend.justification = "bottom"), 0, 0, 1, 1) +
  draw_plot(pK2 + scale_color_viridis(discrete = TRUE) + 
              theme(legend.justification = "top"), 0.5, 0.52, 0.5, 0.4)
plot_grid(pB,p3,pK,ncol=3,labels = c("A","B","C"))
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/OR.RELAX.pdf",width=8,height=5)

```

## plot background Omega
```{r}
backgroundOmegaSF<-data.frame(cbind(allRELAX$ToMean,as.character(allRELAX$sf),as.character(allRELAX$col)))
colnames(backgroundOmegaSF)<-c("T","SF","col")
backgroundOmegaSF$T<-as.numeric(as.character(backgroundOmegaSF$T))
pBsf <- ggplot(backgroundOmegaSF, aes(x=SF, y=T,fill=col)) + 
  geom_violin(position = position_dodge(width =.1),lwd=.1) + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) + 
  scale_y_continuous(name ="background omega") +
  scale_x_discrete(name ="") +
  guides(fill=FALSE)+
  geom_boxplot(width=0.5,outlier.alpha = 0,fill=rgb(1,1,1,.8))
pBsf
```

## plot all results
```{r fig.height=2, fig.width=3}
backgroundKSF<-data.frame(cbind(allRELAX$klog2,as.character(allRELAX$sf),as.character(allRELAX$col)))
colnames(backgroundKSF)<-c("T","SF","col")
backgroundKSF$T<-as.numeric(as.character(backgroundKSF$T))
backgroundKSF$T[!is.finite(backgroundKSF$T)]<- (-80)
backgroundKSF$col<-as.character(backgroundKSF$col)

bksf2 <-
  backgroundKSF %>%
  group_by(SF) %>%
  mutate(outlier = T > median(T) + IQR(T) * 1.5) %>%
  mutate(outlier = T < median(T) - IQR(T) * 1.5) %>%
  ungroup

pKsf <- ggplot(bksf2, aes(x=SF, y=T,fill=col)) + 
  geom_violin(lwd=.1,alpha=.4,fill="white") + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) + 
  scale_y_continuous(name ="log2(k)",limits = c(-90,20)) +
  scale_x_discrete(name ="") +
  guides(fill=FALSE)+
  geom_boxplot(width=.8,outlier.shape = NA, fill=unique(bksf2$col),lwd=.2, varwidth = F) +
  geom_point(data = function(x) dplyr::filter_(x, ~ outlier), position = 'jitter',size=.7,col=rgb(0,0,0,.8),pch=1) +
  theme(legend.position="none")

pKsf
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/OR.sf.RELAX.pdf",width=4,height=6)
#bksf2


```

## plot clades with very low K
```{r}
klogDF<-allRELAX
klogDF$klog2[!is.finite(klogDF$klog2)]<-(-80)
plot(klogDF$klog2,cex=.5,pch=19,col=addalpha(klogDF$col,1),xaxt="n",xlab="",ylab="log2(k)",xlim=c(0,length(klogDF$klog2)+20))
klogDF$index<-1:length(klogDF$OGid)
klogLST<-split(klogDF,f = klogDF$sf)

lowK<-subset(klogDF,klog2 <= -20)
klogLST<-split(klogDF,f = klogDF$sf)

for(i in 1:length(klogLST)){
  par(xpd=NA)
  rect(xleft = klogLST[[i]]$index[1]-.5,xright = klogLST[[i]]$index[length(klogLST[[i]]$index)],ybottom = 14,ytop = -83,border=NA,col=addalpha(klogLST[[i]]$col[1],.3))
  par(xpd=T)
  mtext(text = names(klogLST)[i],side = 3,at=mean(c(klogLST[[i]]$index[1]-.5,klogLST[[i]]$index[length(klogLST[[i]]$index)])),cex=.5)
}
points(klogDF$klog2,cex=.5,pch=19,col=addalpha(klogDF$col,1))


text(x = (lowK$index+10),y=jitter(lowK$klog2),labels = gsub(".*Clade","c",lowK$OGid),cex=.7,srt=45,pos=3)



dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/OR.sf.RELAX.dotplot.pdf",width=4,height=4)

```
## show summary
```{r}
sum(summary(lowK$sf))
summary(lowK$sf)
```

