---
title: "Inquiline Genomics: OR gene tree analysis and plot"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: true
    number_sections: true
    df_print: paged
  pdf_document: default
---
<STYLE TYPE="text/css">
<!--
  td{
    font-family: Arial; 
    font-size: 8pt;
    padding:0px;
    cellpadding="0";
    cellspacing="0"
  }
  th {
    font-family: Arial; 
    font-size: 8pt;
    height: 20px;
    font-weight: bold;
    text-align: right;
    background-color: #ccccff;
  }
  table { 
    border-spacing: 0px;
    border-collapse: collapse;
  }
--->
</STYLE>

# Load Environment
## Load libraries

```{r}
######################
# Load packages
######################
library("ape")
library("Biostrings")
library("ggplot2")
library("ggtree")
library("phytools")
library("treeman")
library(diversitree)
library(phytools)
library(geiger)
library(cowplot)
library(RColorBrewer)
source("~/sciebo/librarySchrader.R")
library(gplots)
library(randomcoloR)
library(data.table)
```

## Load rerooted tree
```{r}
############################################
# Load rerooted FastTree tree
############################################
t1<-read.tree("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/ant.OR.rooted.tre")
tree <- as(t1, 'TreeMan')
#prnt<-getPrnt(tree,c("CcosOr-084","CcosOr-256"))
#prnt<-getPrnt(tree,c("CcosOr-084","CcosOr-156"))
#t1 <- as(getSubtree(tree,prnt),"phylo")
```

## Load blast-based subfamily assignments 
```{r}
############################################
# Load subfam assignments
############################################
subfam<-read.table("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/ORsubfamily.assignment.tsv",sep="\t")

```


# Process tree data
## Generate clades of size 
```{r}

############################################
# Generate clades
############################################
cladesTree<-getCladesofSize(t1,5)

# generate overview dataframe
tipList<-lapply(cladesTree, function(x) x$tip.label)
names(tipList)<-paste("",sprintf("%03d", 1:length(tipList)),sep="")
tipDF<-stack(tipList)
cladeSizes<-lapply(tipList, length)
cladeSizesTab<-do.call(rbind,cladeSizes)
plot(cladeSizesTab,cex=.8,pch=19,col=addalpha(goodCols[1],.8))
abline(h=11,col="red",lwd=2,lty=2)
cladeSummary<-lapply(tipList, function(x) summary(as.factor(substr(x,1,4))))
cladeTable<-plyr::ldply(cladeSummary, rbind)
cladeTable[is.na(cladeTable)] <- 0
dim(cladeTable)

```


## Reassign genes AheyOr-372-NTEfd and AchaOr-402-NCfd to correct SFs
```{r}
tipDF2<-merge(tipDF,subfam,by.x="values",by.y="V1")
colnames(tipDF2)<-c("gene","clade","subfamily")

tmp<-subset(tipDF2,subfamily=="H")
tmp$clade<-as.factor(as.character(tmp$clade))
tmp2<-split(tmp,f=tmp$clade)
# AheyOr-372-NTEfd unique in its clade (c-009)
# All other genes in the clade belong to subfamily E, adjust accordingly
tipDF2[tipDF2$gene=="AheyOr-372-NTEfd",]$subfamily<-"E"


tmp<-subset(tipDF2,subfamily=="G")
tmp$clade<-as.factor(as.character(tmp$clade))
tmp2<-split(tmp,f=tmp$clade)
subset(tipDF2,clade=="C-006")
# AchaOr-402-NCfd	only gene in clade to be assigned to 9E, change accordingly
tipDF2[tipDF2$gene=="AchaOr-402-NCfd",]$subfamily<-"G"

# check if both genes are assigned to G and E
subset(tipDF2,gene=="AchaOr-402-NCfd" | gene == "AheyOr-372-NTEfd")

```


## Add subfamily metadata to tree
```{r}
## https://rawgit.com/valentinitnelav/valentinitnelav.github.io/master/assets/2018-01-07-ggtree/2018-01-07-ggtree.html
# Use groupOTU for taxa classification
#t1 <- ggtree::groupOTU(t1, tipList)
#str(t1)
# check all group labels
#levels(attributes(t1)$group) 
subfamList<-unstack(tipDF2[,c("gene","subfamily")])
t1 <- ggtree::groupOTU(t1, subfamList)
#str(t1)
levels(attributes(t1)$group)

```

## retrieve genes according to clade membership from tree object
```{r}

tipDF2list<-lapply(cladesTree, function(x) subset(tipDF2,gene %in% x$tip.label))
names(cladesTree)<-unlist(lapply(tipDF2list,function(x) x$clade[1]))
```
## Reorder sf attribute (skipped)
```{r eval=FALSE, include=FALSE}
# check all group labels
#levels(attributes(tree_grA)$group) 
#levels(attributes(tree_grA)$group)[1] <- "group_A1"

# Reorder factor levels if needed (this controls the order in the legend)
#attributes(tree_grA)$group <- factor(x = attributes(tree_grA)$group, levels = c("group_A3", "group_A2", "group_A1"))
#+ geom_tiplab(size = 4, aes(angle = angle))
```

 
# Generate trait data for genes & clades
```{r}
############################################
# Generate traits
############################################

smallerCladesTrait<-rep(0,length(t1$tip.label))
names(smallerCladesTrait)<-t1$tip.label

## Tier 1: clade evolution
## level 1: All clades decreased in one parasite
# level 1a: All clades decreased in Parg relative to Ahey
smallerClades1a<-as.numeric(row.names(subset(cladeTable,(Parg<Ahey   & Ahey <= Aech))))
smallerClades1aTree<-cladesTree[smallerClades1a]
smallerClades1aTips<-unlist(lapply(smallerClades1aTree, function(x) x$tip.label))

# level 1b: All decreased in Acha relative to Ahey 
smallerClades1b<-as.numeric(row.names(subset(cladeTable,(Ahey <= Aech & Acha<Ahey))))
smallerClades1bTree<-cladesTree[smallerClades1b]
smallerClades1bTips<-unlist(lapply(smallerClades1bTree, function(x) x$tip.label))

# level 1c: All decreased in Ains relative to Aech
smallerClades1c<-as.numeric(row.names(subset(cladeTable,(Ains<Aech   & Aech <= Ahey))))
smallerClades1cTree<-cladesTree[smallerClades1c]
smallerClades1cTips<-unlist(lapply(smallerClades1cTree, function(x) x$tip.label))

# level 2: Decreases in two parasites
# level 2a: All decreased in Ains & Parg
smallerClades2a<-as.numeric(row.names(subset(cladeTable,(Parg<Ahey) & (Ains<Aech) & (Acha>=Ahey))))
smallerClades2aTree<-cladesTree[smallerClades2a]
smallerClades2aTips<-unlist(lapply(smallerClades2aTree, function(x) x$tip.label))

# level 2b: All decreased in Ains & Acha
smallerClades2b<-as.numeric(row.names(subset(cladeTable,(Acha<Ahey) & (Ains<Aech) & (Parg>=Ahey))))
smallerClades2bTree<-cladesTree[smallerClades2b]
smallerClades2bTips<-unlist(lapply(smallerClades2bTree, function(x) x$tip.label))

# level 2c: All decreased in Parg & Acha (& Ahey >=Aech)
smallerClades2c<-as.numeric(row.names(subset(cladeTable,(Acha<Ahey) & (Parg<Ahey) & (Ahey<=Aech))))
smallerClades2cTree<-cladesTree[smallerClades2c]
smallerClades2cTips<-unlist(lapply(smallerClades2cTree, function(x) x$tip.label))

# level 3:  All clades decreased in all 3 parasites
smallerClades3<-as.numeric(row.names(subset(cladeTable,Parg<Ahey & Acha<Ahey & Ains<Aech)))
smallerClades3Tree<-cladesTree[smallerClades3]
smallerClades3Tips<-unlist(lapply(smallerClades3Tree, function(x) x$tip.label))



# finalize trait data frame
smallerCladesTrait[smallerClades1aTips]<-"Parg"
smallerCladesTrait[smallerClades1bTips]<-"Acha"
smallerCladesTrait[smallerClades1cTips]<-"Ains"
smallerCladesTrait[smallerClades2aTips]<-"Parg/Ains"
smallerCladesTrait[smallerClades2bTips]<-"Acha/Ains"
smallerCladesTrait[smallerClades2cTips]<-"Parg/Acha"
smallerCladesTrait[smallerClades3Tips]<-"Parg/Ains/Acha"

smallerCladesTrait<-smallerCladesTrait[t1$tip.label]
smallerCladesTrait<-as.factor(smallerCladesTrait)

## Tier3: All increased in leaf-cutters relative to outgroups

lcCladesTrait<-rep("0",length(t1$tip.label))
names(lcCladesTrait)<-t1$tip.label

# level 1: All increased in leaf-cutters relative to outgroups
rmedianLC<-apply(cbind(cladeTable$Acep,cladeTable$Acol,cladeTable$Aech,cladeTable$Ahey),1, median, na.rm = TRUE)
rmedianOut<-apply(cbind(cladeTable$Tzet,cladeTable$Tsep,cladeTable$Ccos,cladeTable$Tcor),1, median, na.rm = TRUE)
lcCladesTrees<-cladesTree[rmedianLC>rmedianOut]
lcCladesTips<-unlist(lapply(lcCladesTrees, function(x) x$tip.label))
lcCladesTrait[lcCladesTips]<-"gains in LC"

plot(rmedianLC,rmedianOut)

# finalize trait data frame
lcCladesTrait<-lcCladesTrait[t1$tip.label]
lcCladesTrait<-as.factor(lcCladesTrait)

## Tier 2: subfamilies
# level 1:  All clades decreased in all 3 parasites
subfamTrait<-tipDF2$clade
names(subfamTrait)<-tipDF2$gene
subfamTrait<-subfamTrait[t1$tip.label]
```

## Select certain tips based on traits
```{r}

### Find clades expanding in leafcutters and then shrinking in parasites
# generate one element containing all shrinking clades 
smallerCladesTrees<-modifyList(smallerClades1aTree,smallerClades1bTree) %>% modifyList(.,smallerClades1cTree) %>% modifyList(.,smallerClades2aTree)  %>% modifyList(.,smallerClades2bTree) %>% modifyList(.,smallerClades2cTree) %>% modifyList(.,smallerClades3Tree)

# extract first tip.label for each LC-growing clade and each shrinking clade
lcCladesFirstTip<-unlist(lapply(lcCladesTrees, function(x) x$tip.label[1]))
smallerCladesFirstTip<-unlist(lapply(smallerCladesTrees, function(x) x$tip.label[1]))

# extract first tip.label for Parg shrinking clades
PargCladesFirstTip<-unlist(lapply(smallerClades1aTree, function(x) x$tip.label[1]))

# get all those clades that share the first tip between the LC-growing and parasite shrinking clade
lc2PARG<-which(lcCladesFirstTip %in% PargCladesFirstTip)
lc2PARGtree<-lcCladesTrees[lc2PARG]

lc2smaller<-which(lcCladesFirstTip %in% smallerCladesFirstTip)
lc2smallerTree<-lcCladesTrees[lc2smaller]

```

## Plot some clades if desired
```{r}

class(lc2smallerTree) <- "multiPhylo"
ggtree(lc2smallerTree,ladderize = T,right = T) + facet_wrap(~.id, scale="free", ncol=8,nrow = 5) + ggtitle("All clades expanding in leaf-cutters and shrinking in parasites.") +   geom_tiplab(cex=1) 
```

## Plot heatmap of gene copies in clades vs species
```{r}
cladeMatrixLC<-cladeTable[,c("Aech","Ains","Parg","Ahey","Acha","Acol","Acep")]
color.palette  <- colorRampPalette(c("gray90","blue"))(n=10)
heatmap.2(as.matrix(cladeMatrixLC), col = color.palette)
```

## create color vectors for all trait data
```{r}
n <- length(levels(subfamTrait))
n2 <- length(levels(smallerCladesTrait))
n3 <- length(levels(lcCladesTrait))
n4 <- length(levels(attributes(t1)$group))
n5<- length(levels(tipDF2$subfamily))
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]

# subfamilies or clades
#col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))[1:n]
#col_vector = c("black",distinctColorPalette(n))
#col_vector = rep(c("gray60","gray80"),length=n)
col_vector <- rep(rgb(0,0,0,0),n)
# parasite losses
col_vector2 = c("gray90",unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))[(n+2):(n+n2)])
col_vector2 <- addalpha(c("gray90",goodCols[1],goodCols[2],goodCols[3],goodCols[5]),.5)

fc<-colorRampPalette(c("#FFCCCC", goodCols[8]))
col_vector2 <- c("gray90",fc(n2-1))
#plot(1:10,1:10,pch=19,col=col_vector2,cex=5)

# LC increase
#col_vector3 = c("gray80",unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))[(n+n2+2):(n+n2+n3)])
col_vector3 <-c (rgb(0,0,0,0),addalpha(goodCols2[6],0.6))
#plot(1:10,1:10,pch=19,col=col_vector3,cex=5)

# branch coloring
#col_vector4 = c("black",unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))[1:length(levels(attributes(t1)$group))])
#col_vector4 <- c("black",distinctColorPalette(n4))
#col_vector4 <- c("black",gray.colors(n4))
col_vector4 <- rep("black",(n4+1))

# subfamily coloring
col_vector5 <- distinctColorPalette(n5)
```

## prepare tree with 3 outside heatmaps
```{r}
### 3 outline circles

circ <- ggtree(t1,size=.2, layout = "circular",ladderize = T,right = T,mapping = aes(color = group)) + scale_color_manual(name = 'group',values = col_vector4)

# 1st tier: Subfamilies
p1 <- gheatmap(circ, data.frame(subfamTrait), offset=0, width=.02,
               colnames_angle=90, colnames_offset_y = .25,color = NA,colnames = F)
# 2nd tier: Clades changing in parasites
p2 <- gheatmap(p1, data.frame(smallerCladesTrait), offset=.15, width=.02,
               colnames_angle=90, colnames_offset_y = .25,color = NA,colnames = F)
# third tier: Clades larger in leaf-cutters than in outgroups
p3 <- gheatmap(p2, data.frame(lcCladesTrait), offset=0, width=.02,
               colnames_angle=90, colnames_offset_y = .25,color = NA,colnames = F)

# third tier added
col <- c(col_vector, col_vector2,col_vector3)
names(col) = c(levels(subfamTrait), levels(smallerCladesTrait),levels(lcCladesTrait))

# changed to third tier
pp <- p3 + scale_fill_manual(values=col)


p1x <- p1 + scale_fill_manual(values=col)
p2x <- gheatmap(circ, data.frame(smallerCladesTrait), offset=.15, width=.02,color = NA) +
    scale_fill_manual(values=col) 

# third tier 
p3x <- gheatmap(circ, data.frame(lcCladesTrait), offset=0, width=.02,color = NA) +
    scale_fill_manual(values=col) 

#leg1 <- get_legend(p1x)
leg2 <- get_legend(p2x)
leg3 <- get_legend(p3x)

pp <- pp + theme(legend.position="none") 
```

## plot simple tree
```{r}

#pdf("/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/ggORtreeClean.pdf",width=20,height=20)
#plot_grid(pp, leg1, leg2, leg3, ncol=4, rel_widths=c(1, .1, .1,.1))

plot_grid(leg2, leg3, ncol=2)
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/ggORtreeClean.legend.pdf",width=10,height=10)
#dev.off()
#pp
#dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/ggORtreeClean.pdf",width=10,height=10)

```

## add clade info to clades in ggtree object
```{r}
pp2 <- pp %<+% tipDF2
tree_dt <- data.table(pp2$data)


# select only the tip labels and order by coord y
tree_dt <- tree_dt[isTip == TRUE][order(y)]

# Make table with y cords for each group in consecutive order; 
# this helps for drawing & labeling segments.
# Note the usage of "rleid" function, which is a grouping ID generator,
# needed because consecutive rows of an identical reoccurring group must form a unique group.
coord_groups <- tree_dt[, .(y1 = y[1],
                            y2 = y[.N],
                            angle = mean(angle),
                            n = .N), # optional - helps with counting
                        by = .(clade, 
                               id_gr = rleid(clade, 
                                             prefix = "cld"))]
coord_groups


# Compute the middle y - will be used for placing the group label;
# similarly the mean angle was computed above already. 
coord_groups[, y_mid := rowMeans(.SD), .SDcols = c("y1", "y2")]

# For one-record groups where y1=y2, adjust their y coordinates so that a segment gets drawn;
# If not, no segment get drawn for such cases.
# To force a segment add and subtract a small amount (try & error until seems ok). 
# Prefer a smaller values since with bigger ones you risk to exaggerate the segments.
coord_groups[, y1_adj := ifelse(y1 == y2, y1 - 0.1, y1)]
coord_groups[, y2_adj := ifelse(y1 == y2, y2 + 0.1, y2)]

# Labels need angle adjustment for cases between 90 and 270 dg
coord_groups[, angle_adj := ifelse(angle %between% c(90, 180), 
                                   yes = angle + 180,
                                   no = ifelse(angle > 180 & angle <= 270,
                                               yes = angle - 180,
                                               no = angle))]

# Labels with angles between 90 and 270 dg
# need change of horizontal adjustment argument from 0 to 1.
coord_groups[, hjust_adj := ifelse(angle %between% c(90, 270), yes = 1L, no = 0L)]

# if needed, coloring could be binary 
coord_groups[, col := ifelse(.I%%2, 0.35, 1)]
coord_groups


# Define variable to control x coordinate of segments & labels
my_x <- max(tree_dt$x) + 0.15

tree_labeled <- 
  pp2 + 
  # Add line segments for each group.
  geom_segment(data = coord_groups,
               aes(x = my_x, 
                   y = y1_adj, 
                   xend = my_x, 
                   yend = y2_adj),
               color = rgb(.5,.5,.5,.5),
               lineend = "butt",
               size = 1.2) +
  # Add text group labels at the middle of each segment.
  geom_text(data = coord_groups,
            aes(x = my_x+0.08,
                y = y_mid,
                angle = angle_adj,
                hjust = hjust_adj,
                label = clade),
            vjust = 0.5, 
            size  = 1.2,
            nudge_x = 0.05, # Offsetting label from its default x coordinate.
            color = "black")
```

## add subfamily (here: "group") info to clades in ggtree object
```{r}
coord_groups2 <- tree_dt[, .(y1 = y[1],
                            y2 = y[.N],
                            angle = mean(angle),
                            n = .N), # optional - helps with counting
                        by = .(group, 
                               id_gr = rleid(group, 
                                             prefix = "sf"))]
coord_groups2


# Compute the middle y - will be used for placing the group label;
# similarly the mean angle was computed above already. 
coord_groups2[, y_mid := rowMeans(.SD), .SDcols = c("y1", "y2")]

# For one-record groups where y1=y2, adjust their y coordinates so that a segment gets drawn;
# If not, no segment get drawn for such cases.
# To force a segment add and subtract a small amount (try & error until seems ok). 
# Prefer a smaller values since with bigger ones you risk to exaggerate the segments.
coord_groups2[, y1_adj := ifelse(y1 == y2, y1 - 0.1, y1)]
coord_groups2[, y2_adj := ifelse(y1 == y2, y2 + 0.1, y2)]

# Labels need angle adjustment for cases between 90 and 270 dg
coord_groups2[, angle_adj := ifelse(angle %between% c(90, 180), 
                                   yes = angle + 180,
                                   no = ifelse(angle > 180 & angle <= 270,
                                               yes = angle - 180,
                                               no = angle))]

# Labels with angles between 90 and 270 dg
# need change of horizontal adjustment argument from 0 to 1.
coord_groups2[, hjust_adj := ifelse(angle %between% c(90, 270), yes = 1L, no = 0L)]

# if needed, coloring could be binary 
coord_groups2[, col := ifelse(.I%%2, 0.5, 1)]
coord_groups2

# Define variable to control x coordinate of segments & labels
my_x2 <- max(tree_dt$x) + .55

tree_labeled2 <- 
  tree_labeled + 
  # Add line segments for each group.
  geom_segment(data = coord_groups2,
               aes(x = my_x2, 
                   y = y1_adj, 
                   xend = my_x2, 
                   yend = y2_adj),
               color = col_vector5,
               lineend = "butt",
               size = 2) +
  # Add text group labels at the middle of each segment.
  geom_text(data = coord_groups2,
            aes(x = my_x2+0.1,
                y = y_mid,
                angle = angle_adj,
                hjust = hjust_adj,
                label = group),
            vjust = 0.5, 
            size  = 2,
            nudge_x = 0.05, # Offsetting label from its default x coordinate.
            color = "black")
```

# Plot final tree
```{r}
tree_labeled2
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/OR.labeled.large.pdf",width=10,height=10)
dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/OR.labeled.small.pdf",width=5.5,height=5.5)
```

# Save relevant objects
```{r}
dfCT<-data.frame(lcCladesTrait)
tipDF3<-merge(tipDF2,dfCT,by.x="gene",by.y="row.names")
dfSC<-data.frame(smallerCladesTrait)
tipDF4<-merge(tipDF3,dfSC,by.x="gene",by.y="row.names")

# save some things
save(cladesTree,file="/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/cladesTree.Robject")
save(tree_dt,file="/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/tree_dt.Robject")
write.table(tipDF4,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/tipsDF4.tsv",quote=F,sep="\t",row.names = F)
write.table(tree_dt,"/Users/lukas/sciebo/inquilineGenomics18/ORevolution/results/tree/tree_dt.tsv",quote=F,sep="\t",row.names = F)
```

