---
title: "Genome evolution of inquilines"
output:
  html_document:
    df_print: paged
---


```{r}
source("~/sciebo/librarySchrader.R")
library(dplyr)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(gplots)
```

###Load and prepare the data from whole genome alignments.
```{r include=FALSE}
rawDataIns<-list()
rawDataDel<-list()
data<-list()
data2<-list()
resultFolder<-"/Users/lukas/sciebo/inquilineGenomics18/WGA/results/"
#selection<- c("ACHA","PARG","AHEY","AINS","AECH","Anc09","ACOL","ACEP","Anc07","Anc08","Anc05","Anc04","Anc06")
selection<- c("ACHA","PARG","AHEY","AINS","AECH","Anc09","ACOL","ACEP","Anc07","Anc08","Anc05","Anc04")
for (species in selection){

  rawDataIns[[species]]<-read.csv(paste(resultFolder,species,".ins.bed",sep=""),sep="\t",comment.char = "#",F)
  rawDataDel[[species]]<-read.csv(paste(resultFolder,species,".del.bed",sep=""),sep="\t",comment.char = "#",F)
  colnames(rawDataIns[[species]])<-c("Sequence", "Start", "End","MutationID","ParentGenome","ChildGenome")
  colnames(rawDataDel[[species]])<-c("Sequence", "Start", "End","MutationID","ParentGenome","ChildGenome")
  data[[species]]<-rbind(rawDataDel[[species]],rawDataIns[[species]])
  data2[[species]]<-split(data[[species]],f=data[[species]]$MutationID)
}

```
### Retrieve progressive Cactus mutations.tsv
```{r}
 muts<-read.csv("/Users/lukas/sciebo/inquilineGenomics18/WGA/results/mutations.attines.tsv",sep=",")
```

*"ACHA","PARG","AHEY","AINS","AECH","Anc09","ACOL","ACEP","Anc07","Anc08","Anc05","Anc04","Anc06"*

 P=Transposition
 I=Insertion
 D=Deletion
 V=Inversion
 GI(D)=GapInsertion(GapDeletion)
 U=Duplication
 DB=Deletion Breakpoint GDB=Gap Deletion Breakpoint

###Create subsets for each mutation type and each species/branch.
```{r include=FALSE}
# P=Transposition
# I=Insertion
# D=Deletion
# V=Inversion
# GI(D)=GapInsertion(GapDeletion)
# U=Duplication
# DB=Deletion Breakpoint GDB=Gap Deletion Breakpoint

P<-list()
D<-list()
V<-list()
U<-list()
I<-list()
Psummary<-list()
Dsummary<-list()
Vsummary<-list()
Usummary<-list()
Isummary<-list()

PRanges<-c(0,100,1000,2000,5000,10000,20000,30000,40000,500000,1000000,2000000,2000000000)
DRanges<-c(0,10,50,100,500,1000,1500,2000,2500,5000,10000,20000,30000,40000)
VRanges<-c(0,10,50,100,500,1000,1500,2000,2500,5000,10000,20000,2000000000)
URanges<-c(0,10,50,100,500,1000,1500,2000,2500,5000,10000,20000,30000,40000)
IRanges<-c(0,10,50,100,500,1000,1500,2000,2500,5000,10000,20000,30000,40000)

for (species in selection){
  P[[species]]<-data2[[species]]$P
  P[[species]]$length<-P[[species]]$End-P[[species]]$Start
  D[[species]]<-data2[[species]]$D
  D[[species]]$length<-D[[species]]$End-D[[species]]$Start
  V[[species]]<-data2[[species]]$V
  V[[species]]$length<-V[[species]]$End-V[[species]]$Start
  U[[species]]<-data2[[species]]$U
  U[[species]]$length<-U[[species]]$End-U[[species]]$Start
  I[[species]]<-data2[[species]]$I
  I[[species]]$length<-I[[species]]$End-I[[species]]$Start


  P[[species]]$group<-cut(P[[species]]$length, breaks = PRanges, right = TRUE)
  Psummary[[species]]<-P[[species]] %>%
    group_by(group) %>%
    count(group)
  Psummary[[species]]$species<-species
  Psummary[[species]]$MY<-Psummary[[species]]$n/(muts$BranchLength[muts$GenomeName==species]*100)

  D[[species]]$group<-cut(D[[species]]$length, breaks = DRanges, right = TRUE)
  Dsummary[[species]]<-D[[species]] %>%
    group_by(group) %>%
    count(group)
  Dsummary[[species]]$species<-species
  Dsummary[[species]]$MY<-Dsummary[[species]]$n/(muts$BranchLength[muts$GenomeName==species]*100)

  V[[species]]$group<-cut(V[[species]]$length, breaks = VRanges, right = TRUE)
  Vsummary[[species]]<-V[[species]] %>%
    group_by(group) %>%
    count(group)
  Vsummary[[species]]$species<-species
  Vsummary[[species]]$MY<-Vsummary[[species]]$n/(muts$BranchLength[muts$GenomeName==species]*100)

  U[[species]]$group<-cut(U[[species]]$length, breaks = URanges, right = TRUE)
  Usummary[[species]]<-U[[species]] %>%
    group_by(group) %>%
    count(group)
  Usummary[[species]]$species<-species
  Usummary[[species]]$MY<-Usummary[[species]]$n/(muts$BranchLength[muts$GenomeName==species]*100)


  I[[species]]$group<-cut(I[[species]]$length, breaks = IRanges, right = TRUE)
  Isummary[[species]]<-I[[species]] %>%
    group_by(group) %>%
    count(group)
  Isummary[[species]]$species<-species
  Isummary[[species]]$MY<-Isummary[[species]]$n/(muts$BranchLength[muts$GenomeName==species]*100)
}
```

### Calculate and plot

#### V=Inversions
```{r echo=FALSE}

##################################################################
# V=Inversions
##################################################################

Vall<-do.call("rbind", Vsummary)
Vall$species<-as.factor(Vall$species)
Vall$species<-(factor(Vall$species,levels(Vall$species)[c(7,9,1,3,8,10,4,6,11,5,12,2,13)]))
# rename labels for ranges
levels(Vall$group)[levels(Vall$group)=="(0,10]"]<-" 0.01 kb"
levels(Vall$group)[levels(Vall$group)=="(10,50]"]<-" 0.01 kb"
levels(Vall$group)[levels(Vall$group)=="(50,100]"]<-" 0.05-0.1 kb"
levels(Vall$group)[levels(Vall$group)=="(100,500]"]<-" 0.1-0.5 kb"
levels(Vall$group)[levels(Vall$group)=="(500,1e+03]"]<-" 0.5-1.0 kb"
levels(Vall$group)[levels(Vall$group)=="(1e+03,1.5e+03]"]<-" 1.0-1.5 kb"
levels(Vall$group)[levels(Vall$group)=="(1.5e+03,2e+03]"]<-" 1.5-2.0 kb"
levels(Vall$group)[levels(Vall$group)=="(2e+03,2.5e+03]"]<-" 2.0-2.5 kb"
levels(Vall$group)[levels(Vall$group)=="(2.5e+03,5e+03]"]<-" 2.5-5.0 kb"
levels(Vall$group)[levels(Vall$group)=="(5e+03,1e+04]"]<-" 5.0-10 kb"
levels(Vall$group)[levels(Vall$group)=="(1e+04,2e+04]"]<-" 10-20 kb"
levels(Vall$group)[levels(Vall$group)=="(2e+04,2e+09]"]<-" 20 kb"

Vall$group<-as.factor(Vall$group)
#levels(Vall$group)<-levels(Vall$group)[c(9,1:8)]
#levels(Vall)<-rev(levels(Vall))

theme_set(theme_pubclean())

cols <- colorRampPalette(brewer.pal(8,"Blues"))(length(levels(Vall$group)))
ggplot(Vall, aes(x = species, y = MY)) +
  geom_bar(aes(color = Vall$group, fill = Vall$group), stat = "identity") +
  scale_color_manual(values = rep("gray40",20))  + theme(legend.position="right") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = cols) +
  ylab("Inversion Events per MY") +
  xlab("Branch")

```

# D=Deletions
```{r echo=FALSE}

##################################################################
# Deletions
##################################################################

Dall<-do.call("rbind", Dsummary)
Dall$species<-as.factor(Dall$species)
Dall$species<-(factor(Dall$species,levels(Dall$species)[c(7,9,1,3,8,10,4,6,11,5,12,2,13)]))


#c(0,50,100,500,1000,1500,2000,2500,5000,10000,20000,30000,40000)
# rename labels for ranges
levels(Dall$group)[levels(Dall$group)=="(0,10]"]<-" 0.01 kb"
levels(Dall$group)[levels(Dall$group)=="(10,50]"]<-" 0.01-0.05 kb"
levels(Dall$group)[levels(Dall$group)=="(50,100]"]<-" 0.05-0.1 kb"
levels(Dall$group)[levels(Dall$group)=="(100,500]"]<-" 0.1-0.5 kb"
levels(Dall$group)[levels(Dall$group)=="(500,1e+03]"]<-" 0.5-1.0 kb"
levels(Dall$group)[levels(Dall$group)=="(1e+03,1.5e+03]"]<-" 1.0-1.5 kb"
levels(Dall$group)[levels(Dall$group)=="(1.5e+03,2e+03]"]<-" 1.5-2.0 kb"
levels(Dall$group)[levels(Dall$group)=="(2e+03,2.5e+03]"]<-" >2.0 kb"

Dall$group<-as.factor(Dall$group)
#levels(Dall$group)<-levels(Dall$group)[c(9,1:8)]
#levels(Dall)<-rev(levels(Dall))

library(dplyr)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)

theme_set(theme_pubclean())

cols <- colorRampPalette(brewer.pal(8,"Blues"))(length(levels(Dall$group)))
ggplot(Dall, aes(x = species, y = MY)) +
  geom_bar(aes(color = Dall$group, fill = Dall$group), stat = "identity") +
  scale_color_manual(values = rep("gray40",30))  + theme(legend.position="right") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = cols) +
  ylab("Deletion Events per MY") +
  xlab("Branch")

```

# P=Transpositions
```{r echo=FALSE}

##################################################################
# P=Transpositions
##################################################################

Pall<-do.call("rbind", Psummary)
Pall$species<-as.factor(Pall$species)
Pall$species<-(factor(Pall$species,levels(Pall$species)[c(7,9,1,3,8,10,4,6,11,5,12,2,13)]))

# rename labels for ranges
levels(Pall$group)[levels(Pall$group)=="(0,100]"]<-" 0.1 kb"
levels(Pall$group)[levels(Pall$group)=="(100,1e+03]"]<-" 0.1-1.0 kb"
levels(Pall$group)[levels(Pall$group)=="(1e+03,2e+03]"]<-" 1.0-2.0 kb"
levels(Pall$group)[levels(Pall$group)=="(2e+03,5e+03]"]<-" 2.0-5.0 kb"
levels(Pall$group)[levels(Pall$group)=="(5e+03,1e+04]"]<-" 5.0-10 kb"
levels(Pall$group)[levels(Pall$group)=="(1e+04,2e+04]"]<-" 10-20 kb"
levels(Pall$group)[levels(Pall$group)=="(2e+04,3e+04]"]<-" 20-30 kb"
levels(Pall$group)[levels(Pall$group)=="(3e+04,4e+04]"]<-" 30-40 kb"
levels(Pall$group)[levels(Pall$group)=="(4e+04,5e+05]"]<-" >40 kb"



Pall$group<-as.factor(Pall$group)
#levels(Pall$group)<-levels(Pall$group)[c(9,1:8)]
#levels(Pall)<-rev(levels(Pall))

library(dplyr)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)

theme_set(theme_pubclean())

cols <- colorRampPalette(brewer.pal(8,"Blues"))(length(levels(Pall$group)))
ggplot(Pall, aes(x = species, y = MY)) +
  geom_bar(aes(color = Pall$group, fill = Pall$group), stat = "identity") +
  scale_color_manual(values = rep("gray40",30))  + theme(legend.position="right") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = cols) +
  ylab("Transposition Events per MY") +
  xlab("Branch")
```

# U=Duplications
```{r echo=FALSE}

##################################################################
# U=Duplication
##################################################################

Uall<-do.call("rbind", Usummary)
Uall$species<-as.factor(Uall$species)
Uall$species<-(factor(Uall$species,levels(Uall$species)[c(7,9,1,3,8,10,4,6,11,5,12,2,13)]))


# rename labels for ranges
levels(Uall$group)[levels(Uall$group)=="(0,10]"]<-" 0.01 kb"
levels(Uall$group)[levels(Uall$group)=="(10,50]"]<-" 0.01-0.05 kb"
levels(Uall$group)[levels(Uall$group)=="(50,100]"]<-" 0.05-0.1 kb"
levels(Uall$group)[levels(Uall$group)=="(100,500]"]<-" 0.1-0.5 kb"
levels(Uall$group)[levels(Uall$group)=="(500,1e+03]"]<-" 0.5-1.0 kb"
levels(Uall$group)[levels(Uall$group)=="(1e+03,1.5e+03]"]<-" 1.0-1.5 kb"
levels(Uall$group)[levels(Uall$group)=="(1.5e+03,2e+03]"]<-" 1.5-2.0 kb"
levels(Uall$group)[levels(Uall$group)=="(2e+03,2.5e+03]"]<-" 2.0-2.5 kb"
levels(Uall$group)[levels(Uall$group)=="(2.5e+03,5e+03]"]<-" 2.5-5.0 kb"
levels(Uall$group)[levels(Uall$group)=="(5e+03,1e+04]"]<-" 5.0-10 kb"
levels(Uall$group)[levels(Uall$group)=="(1e+04,2e+04]"]<-" 10-20 kb"
levels(Uall$group)[levels(Uall$group)=="(2e+04,3e+04]"]<-" 20-30 kb"
levels(Uall$group)[levels(Uall$group)=="(3e+04,4e+04]"]<-" 30-40 kb"
levels(Uall$group)[levels(Uall$group)=="(4e+04,5e+05]"]<-" >40 kb"


Uall$group<-as.factor(Uall$group)
#levels(Uall$group)<-levels(Uall$group)[c(9,1:8)]
#levels(Uall)<-rev(levels(Uall))

library(dplyr)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)

theme_set(theme_pubclean())

cols <- colorRampPalette(brewer.pal(8,"Blues"))(length(levels(Uall$group)))
ggplot(Uall, aes(x = species, y = MY)) +
  geom_bar(aes(color = Uall$group, fill = Uall$group), stat = "identity") +
  scale_color_manual(values = rep("gray40",30))  + theme(legend.position="right") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = cols) +
  ylab("Duplication Events per MY") +
  xlab("Branch")

```


# I=Insertions
```{r echo=FALSE}

##################################################################
# I=Insertions
##################################################################

Iall<-do.call("rbind", Isummary)
Iall$species<-as.factor(Iall$species)
Iall$species<-(factor(Iall$species,levels(Iall$species)[c(7,9,1,3,8,10,4,6,11,5,12,2,13)]))


# rename labels for ranges
levels(Iall$group)[levels(Iall$group)=="(0,10]"]<-" 0.01 kb"
levels(Iall$group)[levels(Iall$group)=="(10,50]"]<-" 0.01-0.05 kb"
levels(Iall$group)[levels(Iall$group)=="(50,100]"]<-" 0.05-0.1 kb"
levels(Iall$group)[levels(Iall$group)=="(100,500]"]<-" 0.1-0.5 kb"
levels(Iall$group)[levels(Iall$group)=="(500,1e+03]"]<-" 0.5-1.0 kb"
levels(Iall$group)[levels(Iall$group)=="(1e+03,1.5e+03]"]<-" 1.0-1.5 kb"
levels(Iall$group)[levels(Iall$group)=="(1.5e+03,2e+03]"]<-" 1.5-2.0 kb"
levels(Iall$group)[levels(Iall$group)=="(2e+03,2.5e+03]"]<-" 2.0-2.5 kb"
levels(Iall$group)[levels(Iall$group)=="(2.5e+03,5e+03]"]<-" 2.5-5.0 kb"
levels(Iall$group)[levels(Iall$group)=="(5e+03,1e+04]"]<-" 5.0-10 kb"
levels(Iall$group)[levels(Iall$group)=="(1e+04,2e+04]"]<-" 10-20 kb"
levels(Iall$group)[levels(Iall$group)=="(2e+04,3e+04]"]<-" 20-30 kb"
levels(Iall$group)[levels(Iall$group)=="(3e+04,4e+04]"]<-" 30-40 kb"
levels(Iall$group)[levels(Iall$group)=="(4e+04,5e+05]"]<-" >40 kb"



Iall$group<-as.factor(Iall$group)
#levels(Iall$group)<-levels(Iall$group)[c(9,1:8)]
#levels(Iall)<-rev(levels(Iall))

library(dplyr)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)

theme_set(theme_pubclean())

cols <- colorRampPalette(brewer.pal(8,"Blues"))(length(levels(Iall$group)))
ggplot(Iall, aes(x = species, y = MY)) +
  geom_bar(aes(color = Iall$group, fill = Iall$group), stat = "identity") +
  scale_color_manual(values = rep("gray40",12))  + theme(legend.position="right") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_fill_manual(values = cols) +
  ylab("Insertion Events per MY") +
  xlab("Branch")

```

#### Overall summary
```{r}
sumV<-aggregate(Vall$MY, by=list(species=Vall$species), FUN=sum)
sumP<-aggregate(Pall$MY, by=list(species=Pall$species), FUN=sum)
sumD<-aggregate(Dall$MY, by=list(species=Dall$species), FUN=sum)
sumU<-aggregate(Uall$MY, by=list(species=Uall$species), FUN=sum)
sumI<-aggregate(Iall$MY, by=list(species=Iall$species), FUN=sum)
colnames(sumV)[2]<-"Inv"
colnames(sumP)[2]<-"Tra"
colnames(sumD)[2]<-"Del"
colnames(sumU)[2]<-"Dup"
colnames(sumI)[2]<-"Ins"
all<-merge(sumV,sumP,by="species") %>% merge(.,sumD,by="species") %>% merge(.,sumU,by="species")  %>% merge(.,sumI,by="species") 

m<-as.matrix(all[,-1])
row.names(m)<-all$species
heatmap.2(t(m),trace="none",scale="row", Rowv=F, Colv=T,col=(brewer.pal(8,"Blues")),dendrogram="column")
plot(all$Inv,all$Tra,cex=all$Inv/10,col="gray90",pch=19,xlab="Inversion Events per MY",ylab="Transposition Events per MY")
text(x=all$Inv,y=all$Tra,all$species,col="black",cex=.6)
plot(all$Del,all$Dup+all$Ins,cex=all$Del/1000,col="gray90",pch=19,xlab="Deletion Events per MY",ylab="Insertion and Duplication Events per MY")
text(all$Del,all$Dup+all$Ins,all$species,col="black",cex=.6)

all$total<-rowSums(all[,-1])
barplot(all$total,las=2,names=all$species,xlab="total number of events per MY")
barplot(all$Inv/all$total,las=2,names=all$species,xlab="freaction of Inversions by all events")
barplot(all$Tra/all$total,las=2,names=all$species,xlab="freaction of Transpositions by all events")

```
