---
title: "Synteny Analysis Inquilines"
output:
  html_document:
    df_print: paged
---

### Load libraries
```{r}
library(gplots)
library("RColorBrewer")
source("~/sciebo/librarySchrader.R")
library(coin)
library(ggplot2)
require(reshape2)
library(jmuOutlier)
library(viridis)

```

### Load the data from iadhore output.

Object        | source
------------- | -------------
genes.txt     | /corefac/cse/lukas/inqGen18/synteny/acromyrmex/output2/genes.txt
MP            | /corefac/cse/lukas/inqGen18/synteny/acromyrmex/output2/multiplicon_pairs.txt
segments      | /corefac/cse/lukas/inqGen18/synteny/acromyrmex/output2/segments.txt
```{r}
# load all gene info
Genes<-read.csv("/Users/lukas/sciebo/inquilineGenomics18/synteny/results/output2/genes.txt",sep="\t",T)

# load mutliplicon pairs
MP<-read.csv("/Users/lukas/sciebo/inquilineGenomics18/synteny/results/output2/multiplicon_pairs.txt",sep="\t",T)

# load multiplicon info
segments<-read.csv("/Users/lukas/sciebo/inquilineGenomics18/synteny/results/output2/segments.txt",sep="\t",T)

segments$multiplicon<-as.factor(segments$multiplicon)

# split by multiplicon
segSplit<-split(segments,f=segments$multiplicon)

```

### Prepare a list of all syntenic regions across the 6 analysed genomes
**Takes some minutes***

```{r eval=FALSE, include=FALSE}
f<-list()
for (i in 1:length(segSplit)){
  #define species and scf for multiplicon
  species1<-as.character(segSplit[[i]]$genome[1])
  species2<-as.character(segSplit[[i]]$genome[2])
  scf1<-as.character(segSplit[[i]]$list[1])
  scf2<-as.character(segSplit[[i]]$list[2])
  # select appropriate multiplicon_pair subset
  MP1<-subset(MP,multiplicon==i)
  MP.subset<-MP1[grep(paste(species1,species2,sep="|"),MP1$X,perl=T),]
  
  # find matches between M1 Parg gene and Ahey gene
  Genes1tmp<-subset(Genes,list==scf1 & genome == species1)
  Genes2tmp<-subset(Genes,list==scf2 & genome == species2)
  # retrieve coordinate of first gene from species 1 for a given multiplicon
  firstC1<-subset(Genes1tmp,id==as.character(segSplit[[i]]$first[1]),coordinate)
  # retrieve coordinate of first gene from species 2 for a given multiplicon
  firstC2<-subset(Genes2tmp,id==as.character(segSplit[[i]]$first[2]),coordinate)
  # retrieve coordinate of last gene from species 1 for a given multiplicon
  lastC1<-subset(Genes1tmp,id==as.character(segSplit[[i]]$last[1]),coordinate)
  # retrieve coordinate of first gene from species 2 for a given multiplicon
  lastC2<-subset(Genes2tmp,id==as.character(segSplit[[i]]$last[2]),coordinate)
  
  Genes1<-subset(Genes1tmp,coordinate<=as.numeric(lastC1) &coordinate>=as.numeric(firstC1)  )
  Genes2<-subset(Genes2tmp,coordinate<=as.numeric(lastC2) &coordinate>=as.numeric(firstC2)  )
  colnames(MP.subset)[1]<-"idM"
  a<-merge(Genes1,MP.subset,by.x="id",by.y="X",all.x=T)

  a<-unique(a, by=id)
  a<-a[!(duplicated(a$id)),]
  b<-merge(Genes2,MP.subset,by.x="id",by.y="gene_x", all.x=T)
  b<-b[!(duplicated(b$id)),]
  colnames(a)[1]<-"idA"
  colnames(b)[1]<-"idB"
  final<-merge(a,b,by.x="idA",by.y="X", all=T)
  
  ff<-final[,c(1,3,4,5,13,18,19,20,16)]
  colnames(ff)<-c("G1","S1","C1","St1","G2","S2","C2","St2","G2b")
  ff$C2<-as.numeric(ff$C2)
  ff$C1<-as.numeric(ff$C1)
  f[[i]]<-ff
  names(f)[i]<-paste(species1,species2,sep="-")
}

```
### A plot to explain the logic behind using Kendall's correlation
```{r}

par(mfrow=c(1,2))
c1<-cor.test(c(1:10),c(1:10),method="kendall")
c2<-cor.test(c(1:10),c(1:5,7:11),method="kendall")

plot(c(1:10),c(1:10),cex.axis=.7)
mtext(paste("Kendall's cor.=",c1$estimate,sep=""),3)
plot(c(1:10),c(1:5,9:13),cex.axis=.7)
mtext(paste("Kendall's cor.=",c2$estimate,sep=""),3)

#dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/synteny/results/explanationPlots.pdf",width=5,height=3)
```

### calculate Kendall's rank correlation coefficient for each syntenic region across all genomes.

```{r eval=FALSE, include=FALSE}

lms<-list()
r2<-NA
cors<-list()
kcor<-NA
for (i in 1:length(f)){
  tmp<-f[[i]]
  # calculate Kendall's rank correlation coefficient for each syntenic region across all genomes.
  cors[[i]]<-cor.test(tmp$C1,tmp$C2,method="kendall")
  # retrieve the correlation coefficient
  kcor[i]<-abs(cors[[i]]$estimate)
}

names(kcor)<-names(f)
kcorAll<-kcor
```

**Remove all entries from syntenic regions that contain less than 25 syntenic genes**
```{r eval=FALSE, include=FALSE}
kcor<-kcorAll
# remove all entries from syntenic regions that contain less than 25 syntenic genes
# basic quality control
for (i in 1:length(f)){
  if(sum(f[[i]][,5]==as.character(f[[i]][,9]),na.rm=T)<25){kcor[i]<-NA}
  #if(sum(f[[i]][,5]==as.character(f[[i]][,9]),na.rm=T)>40){kcor[i]<-NA}
}
#save(kcor,file="/Users/lukas/sciebo/inquilineGenomics18/synteny/results/output2.correlations.Rdata")

```

```{r}
#total number of syntenic regions
length(kcor)
# Number of syntenic regions containing more than 25 genes 
sum(!is.na(kcor))

dimsAll<-unlist(lapply(f,nrow))
summary(lm(kcorAll~dimsAll))

#fSet<-f[which(!is.na(kcor))]
#fSet<-f[which(dimsAll>40 & dimsAll < 80 & !is.na(kcor))]
#kcor[dimsAll > 40 & dimsAll < 80  ]<-NA
dims<-unlist(lapply(fSet,nrow))
#summary(lm(kcor[!is.na(kcor)]~dims))
boxplot(dims[grep("PARG-ACOL|ACOL-PARG",names(dims),perl=T)],dims[grep("AHEY-ACOL|ACOL-AHEY",names(dims),perl=T)],dims[grep("AECH-ACOL|ACOL-AECH",names(dims),perl=T)],dims[grep("AINS-ACOL|ACOL-AINS",names(dims),perl=T)],dims[grep("ACHA-ACOL|ACOL-ACHA",names(dims),perl=T)],outline=T)

length(dims[grep("ACOL-AHEY|AHEY-ACOL",names(dims),perl=T)])
length(dims[grep("ACOL-AECH|AECH-ACOL",names(dims),perl=T)])
length(dims[grep("ACOL-PARG|PARG-ACOL",names(dims),perl=T)])
length(dims[grep("ACOL-ACHA|ACHA-ACOL",names(dims),perl=T)])
length(dims[grep("ACOL-AINS|AINS-ACOL",names(dims),perl=T)])


```


```{r}
sl<-which(kcor<0.95 & !is.na(kcor))
fS<-f[sl]

ACOLcor<-levels(as.factor(grep("-ACOL",names(kcor),value=T)))
ACOLcor<-grep("ACOL-ACOL",ACOLcor,invert = T,value=T)


```

```{r}



par(mfrow=c(3,4))
par(mai=c(.6,.6,.1,.1))
par(mgp=c(0,.55,0))

for(i in which(names(fS) %in% ACOLcor)){
  plot(fS[[i]]$C1,fS[[i]]$C2,xaxt="n",yaxt="n",cex=.2,xlab="",ylab="",col=ifelse(fS[[i]]$St1==fS[[i]]$St2,"blue","orange"))
  mtext(paste(substr(fS[[i]]$G1[1],1,4)," ",fS[[i]]$S1[1],sep=""),1,cex=.5)
  mtext(paste(substr(fS[[i]]$G2[1],1,4)," ",fS[[i]]$S2[1],sep=""),2,cex=.5)
  mtext(paste("Kendall's cor.: ", round(kcor[sl[i]],4),sep=""),3,cex=.5)
  # add axes
  xlims<-c(min(fS[[i]]$C1,na.rm=T),max(fS[[i]]$C1,na.rm=T))
  xlabs<-unique(c(as.character(unlist(subset(fS[[i]],C1 == xlims[1],G1))),as.character(unlist(subset(fS[[i]],C1 == xlims[2],G1)))))
  
  ylims<-c(min(fS[[i]]$C2,na.rm=T),max(fS[[i]]$C2,na.rm=T))
  ylabs<-unique(c(as.character(unlist(subset(fS[[i]],C2 == ylims[1],G2))),as.character(unlist(subset(fS[[i]],C2 == ylims[2],G2)))))
  
  axis(1,at=xlims,labels=xlabs,cex.axis=.8,las=2)
  axis(2,at=ylims,labels=ylabs,cex.axis =.8,las=2)
  #points(xlims,ylims,type="l",lty=2,col="red")
  #points(xlims,c(ylims[2],ylims[1]),type="l",lty=2,col="red")
  if(which(fS[[i]]$C2==min(fS[[i]]$C2,na.rm=T))<which(fS[[i]]$C2==max(fS[[i]]$C2,na.rm=T))){
    segments(x0=xlims[1],x1=xlims[2],y0=ylims[1],y1=ylims[2],col="gray80",lty=2)
  }else{
    segments(x0=xlims[1],x1=xlims[2],y0=ylims[2],y1=ylims[1],col="gray80",lty=2)
  }
}
#dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/synteny/results/examplePlots.pdf",width=7,height=5)
```


```{r}
#load(file="/Users/lukas/sciebo/inquilineGenomics18/synteny/results/output2.correlations.Rdata")
```


#### Retrieve all correlation values for each species pair (in list of lists "ksets")
##### calculate matrices mean, median and number of correlations < 1
```{r}
m<-matrix(nrow=6,ncol=6)
colnames(m)<-c("ACHA","PARG","AHEY","AINS","AECH","ACOL")
rownames(m)<-c("ACHA","PARG","AHEY","AINS","AECH","ACOL")
m2<-m
m3<-m
m4<-m
ksets<-list()
for(col in 1:6){
  ksets[[col]]<-list()
  names(ksets)[col]<-colnames(m)[col]
  for (row in 1:6){
    ksets[[col]][[row]]<-c(kcor[grep(paste(colnames(m)[col],rownames(m)[row],sep="-"),names(kcor))],kcor[grep(paste(colnames(m)[row],rownames(m)[col],sep="-"),names(kcor))])
    names(ksets[[col]])[row]<-rownames(m)[row]
    if (col==row){next}
    # median kendall's correlation
    m[row,col]<-median(ksets[[col]][[row]],na.rm=T)
    # mean kendall's correlation
    m2[row,col]<-mean(ksets[[col]][[row]],na.rm=T)
    # percent of kendall's correlation below cutoff
    m3[row,col]<-(sum(ksets[[col]][[row]]<1,na.rm = T)/length(ksets[[1]][[2]]))*100
    #m3[row,col]<-(sum(ksets[[col]][[row]]<1,na.rm = T)/length(ksets[[1]][[2]]))*100
  }
}

library("RColorBrewer")
heatmap.2(m,dendrogram='both', trace='none',col=rev(brewer.pal(9,"YlGnBu")))
heatmap.2(m2,dendrogram='both', trace='none',col=rev(brewer.pal(9,"YlGnBu")))
heatmap.2(m3,trace='none',col=brewer.pal(9,"YlGnBu"),dendrogram='both')
```

```{r fig.height=4, fig.width=3}
par(mai=c(.5,.6,0,0.1))
AcolComparison<-as.data.frame(cbind(unlist(ksets[["ACOL"]])))
AcolComparison$V2<-as.factor(substr(x = rownames(AcolComparison),1,4))
colnames(AcolComparison)<-c("Kendall","species")
AcolComparison<-subset(AcolComparison,species!="ACOL")
AcolComparison$species<-droplevels(AcolComparison$species)
AcolComparison$species<-(factor(AcolComparison$species,levels(AcolComparison$species)[c(2,3,1,4,5)]))
bp<-boxplot(Kendall~species,data=AcolComparison,plot=F)
cols<-addalpha(rev(viridis(12))[as.numeric(cut(bp$stats[3,],breaks = 10))],.9)
boxplot(Kendall~species,data=AcolComparison,outline=F,col=cols,ylim=c(0.987,1.0015),yaxt="n",xaxt="n",ylab=NA,lwd=1.5,frame=F)
par(mgp=c(3,.2,0))
axis(2,at=c(0.987,1),lwd = 1,lwd.ticks = .5,cex.axis=.7,las=2,tck=-0.01)
mtext(2,at=mean(c(0.987,1)),text = expression(tau),cex = 2,line = 1.5,las=2)
axis(1,labels = NA,lwd = 1,lwd.ticks = .5,line = -0.4,tck=-0.03)
par(xpd=NA)
text(y=0.986,x=1:5+0.4,levels(AcolComparison$species),srt=-45,pos=1)
par(xpd=F)
segments(y0=1.0005,y1=1.0005,x0=0.75,x1=2.25,lwd=1.5)
segments(y0=1.0005,y1=1.0005,x0=2.75,x1=5.25,lwd=1.5)

segments(y0=1.0005,y1=1.001,x0=4,x1=4,lwd=1.5)
segments(y0=1.0005,y1=1.001,x0=1.5,x1=1.5,lwd=1.5)
segments(y0=1.001,y1=1.001,x0=1.5,x1=4,lwd=1.5)

val1<-c(ksets[["ACOL"]][["AECH"]],ksets[["ACOL"]][["AHEY"]])
val2<-c(ksets[["ACOL"]][["AINS"]],ksets[["ACOL"]][["ACHA"]],ksets[["ACOL"]][["PARG"]])
fLab<-as.factor(c(rep("H",length(val1)),rep("P",length(val2))))
wt<-wilcox_test(c(val1,val2)~fLab,distribution = approximate(nresample = 100000),alternative="greater")

pt<-perm.test(val1[!is.na(val1)],val2[!is.na(val2)],stat=median)
text(paste("p <",round(pvalue(wt)[1],3)),y=1.0015,x=2.75,font=2)
#dev.print(pdf,"/Users/lukas/sciebo/inquilineGenomics18/synteny/results/output2/synteny.pdf",width=2,height=4)
```


```{r}

wts<-list()
pts<-list()
val1<-ksets[["ACOL"]][["AECH"]]
val2<-ksets[["ACOL"]][["AHEY"]]
fLab<-as.factor(c(rep("H",length(val1)),rep("P",length(val2))))
#wts[["AECH-AHEY"]]<-wilcox_test(c(val1,val2)~f,distribution = approximate(nresample = 10000),alternative="greater")
pts[["AECH-AHEY"]]<-perm.test(val1[!is.na(val1)],val2[!is.na(val2)],stat=median,alternative="greater")

#wilcox.test(ksets[["ACOL"]][["AECH"]],ksets[["ACOL"]][["AHEY"]])

val1<-ksets[["ACOL"]][["AECH"]]
val2<-ksets[["ACOL"]][["AINS"]]
fLab<-as.factor(c(rep("H",length(val1)),rep("P",length(val2))))
#wts[["AECH-AINS"]]<-wilcox_test(c(val1,val2)~f,distribution = approximate(nresample = 10000),alternative="greater")
pts[["AECH-AINS"]]<-perm.test(val1[!is.na(val1)],val2[!is.na(val2)],stat=median,alternative="greater")
#wilcox.test(ksets[["ACOL"]][["AECH"]],ksets[["ACOL"]][["AINS"]])

val1<-ksets[["ACOL"]][["AHEY"]]
val2<-ksets[["ACOL"]][["ACHA"]]
fLab<-as.factor(c(rep("H",length(val1)),rep("P",length(val2))))
#wts[['AHEY-ACHA']]<-wilcox_test(c(val1,val2)~f,distribution = approximate(nresample = 10000),alternative="greater")
pts[['AHEY-ACHA']]<-perm.test(val1[!is.na(val1)],val2[!is.na(val2)],stat=median,alternative="greater")
#wilcox.test(ksets[["ACOL"]][["AHEY"]],ksets[["ACOL"]][["ACHA"]])

val1<-ksets[["ACOL"]][["AHEY"]]
val2<-ksets[["ACOL"]][["PARG"]]
fLab<-as.factor(c(rep("H",length(val1)),rep("P",length(val2))))
#wts[["AHEY-PARG"]]<-wilcox_test(c(val1,val2)~f,distribution = approximate(nresample = 10000),alternative="greater")
pts[["AHEY-PARG"]]<-perm.test(val1[!is.na(val1)],val2[!is.na(val2)],stat=median,alternative="greater")
#wilcox.test(ksets[["ACOL"]][["AHEY"]],ksets[["ACOL"]][["PARG"]])

val1<-ksets[["ACOL"]][["AECH"]]
val2<-ksets[["ACOL"]][["ACHA"]]
fLab<-as.factor(c(rep("H",length(val1)),rep("P",length(val2))))
#wts[["AECH-ACHA"]]<-wilcox_test(c(val1,val2)~f,distribution = approximate(nresample = 10000),alternative="greater")
pts[["AECH-ACHA"]]<-perm.test(val1[!is.na(val1)],val2[!is.na(val2)],stat=median,,alternative="greater")
#wilcox.test(ksets[["ACOL"]][["AECH"]],ksets[["ACOL"]][["ACHA"]])

```

